---
title: "Theory, computation, and validation of SPI and SPEI with hydroTSM"
author:
- Mauricio Zambrano-Bigiarini
date: "version 1.1, 27-Jul-2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
  html_document:
    df_print: paged
    number_sections: yes
  pdf_document: 
    number_sections: true
    fig_caption: true
vignette: |
  %\VignetteIndexEntry{Theory, computation, and validation of SPI and SPEI with hydroTSM}
  %\VignetteKeyword{hydrology}
  %\VignetteKeyword{drought}
  %\VignetteKeyword{SPI}
  %\VignetteKeyword{SPEI}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.align="center",
  fig.width=7.2,
  fig.height=4.8
)
options(width=90)
```

# Part I: Theoretical background

## What SPI and SPEI measure

The Standardized Precipitation Index (SPI) describes how unusual an accumulated
precipitation amount is relative to the precipitation climatology for the same
location and calendar month. It was introduced as a probability-based,
multiscalar index: only precipitation is required, and the accumulation period
is selected to represent the memory of the system being studied (McKee et al.,
1993; Guttman, 1998; Hayes et al., 1999).

The Standardized Precipitation-Evapotranspiration Index (SPEI) applies the same
standardisation principle to climatic water balance,

$$D_t = P_t - PET_t,$$

where \(P_t\) is precipitation and \(PET_t\) is potential evapotranspiration.
Consequently, SPEI responds both to water supply and to atmospheric evaporative
demand. This makes it useful when temperature variability or long-term warming
affects drought severity, but also makes the result sensitive to the selected
PET method and the quality of its meteorological inputs (Vicente-Serrano et al.,
2010; Beguería et al., 2014).

For either index, let \(X_t\) denote monthly precipitation for SPI or monthly
water balance for SPEI. A backward-looking value at scale \(k\) is

$$Y_{t,k} = \sum_{j=0}^{k-1} w_j X_{t-j},$$

where \(w_j\) are the kernel weights. A separate probability distribution
\(F_m\) is fitted to \(Y_{t,k}\) for each calendar month \(m\), because a
January water balance should be compared with other Januaries rather than with
months from a different season. The fitted non-exceedance probability is then
mapped to a standard-normal variate:

$$I_{t,k} = \Phi^{-1}\{H_m(Y_{t,k})\},$$

where \(\Phi^{-1}\) is the standard-normal quantile function and \(H_m\) is
either the fitted continuous distribution or, for zero precipitation, a mixed
discrete-continuous distribution. If the probability model is well calibrated,
the resulting index has an approximately standard-normal interpretation:
zero is near the climatological median, negative values are drier than normal,
and positive values are wetter than normal. Standard-normal thresholds of
-1, -1.5, and -2 correspond approximately to lower-tail probabilities of
15.9%, 6.7%, and 2.3%, respectively.

SPI and SPEI describe statistical rarity, not drought impact by themselves.
The same index value can have different consequences depending on season,
exposure, water storage, crop stage, and antecedent conditions.

## Temporal scales and system memory

The `scale` is the number of current and preceding months contributing to each
index value. It is not the duration of an identified drought event. A drought
event may persist for many index values, and neighbouring values are correlated
because their accumulation windows overlap.

Common interpretations are:

- 1 month: immediate precipitation or water-balance anomaly and rapidly
  developing meteorological drought;
- 3 months: short seasonal conditions, soil-moisture stress, and rain-fed
  agricultural response;
- 6 months: seasonal water availability and slower agricultural or
  surface-water response;
- 12 months: annual water balance, streamflow, reservoir, and groundwater
  signals; and
- 24 months or longer: persistent hydrological storage deficits.

These are guidelines, not universal definitions. The appropriate scale should
be selected against the response time of the variable or impact of interest.
Multiscalar interpretation is one of the principal advantages of both indices
(Guttman, 1998; Vicente-Serrano et al., 2010). Comparing several scales is often
more informative than choosing a single scale in advance.

## Why distribution choice matters

The probability model determines the percentile assigned to each accumulated
value and therefore directly controls the final index, especially in the
tails. A poor distribution can exaggerate or suppress drought severity even
when the fitted mean appears reasonable (Stagge et al., 2015). Distribution
choice should respect both the support of the data and the empirical shape:

- SPI precipitation is non-negative, usually right-skewed at short scales, and
  may contain an exact probability mass at zero. Skewness commonly decreases
  with longer accumulation.
- SPEI water balance can be negative or positive, so distributions with a
  location parameter and support extending below zero are generally required.

Stagge et al. (2015) evaluated seven SPI candidates (Gamma, Gumbel, logistic,
log-logistic, lognormal, normal, and Weibull) and four SPEI candidates
(generalized logistic, GEV, normal, and Pearson type III). They found the
two-parameter Gamma to be a strong general SPI choice in their European
experiment and recommended GEV for SPEI there. The original SPEI formulation
used the three-parameter log-logistic distribution, which is represented here
by the equivalent generalized-logistic formulation (Vicente-Serrano et al.,
2010). These findings should not be treated as universal winners: performance
can change with climate, record length, calendar month, accumulation scale, and
observational or modelled data (Pieper et al., 2020).

A defensible selection checks every calendar month and scale, with particular
attention to the dry tail. Useful diagnostics include Q-Q or P-P plots,
empirical versus theoretical drought-category frequencies, tail-sensitive
goodness-of-fit statistics, and validation outside the calibration period.
Goodness-of-fit critical values must account for estimated parameters; applying
an ordinary Kolmogorov-Smirnov reference distribution after fitting parameters
to the same sample is not valid without adjustment or resampling.

## Why parameter-estimation method matters

Distribution family and parameter estimator are separate decisions. Two
implementations can use the same named distribution but produce different
indices because their parameter estimates differ.

- Maximum likelihood (`fit="max-lik"`) selects parameters that maximise the
  probability of the calibration sample. Under a correct model and standard
  regularity conditions it is statistically efficient, but it can be sensitive
  to starting values, outliers, parameter boundaries, and small samples.
- Unbiased probability-weighted moments (`fit="ub-pwm"`) estimate moments that
  combine observations with their non-exceedance probabilities using unbiased
  order-statistic weights. PWMs and the related L-moments provide stable
  estimators for many skewed hydrological distributions (Greenwood et al.,
  1979; Hosking, 1990). Beguería et al. (2014) recommend unbiased PWM for the
  SPEI log-logistic model.
- Plotting-position PWM (`fit="pp-pwm"`) replaces the unknown probabilities by
  empirical plotting positions. hydroTSM uses
  \(F_i=(i-0.35)/n\), matching the constants used by the SPEI package. This was
  used in the original SPEI formulation, but it is not identical to unbiased
  PWM and can differ appreciably in the short monthly calibration samples.

There is no estimator that compensates for a badly chosen distribution.
Selection should consider convergence, plausibility of parameters, fit in the
tails, and sensitivity of the resulting drought classes. Comparisons between
packages are fair only when distribution, estimator, accumulation, reference
period, zero treatment, and scaling conventions are equivalent.

## Theoretical guide to the function arguments

The arguments below separate scientific choices from numerical and output
controls. Defaults are intentionally stated because hydroTSM defaults do not
always match those of other R packages.

### Data, accumulation, and calibration

```{r theoretical-data-arguments, echo=FALSE}
data.arguments <- data.frame(
  argument=c(
    "x", "scale", "kernel", "ref.start, ref.end",
    "zero.threshold", "p0", "p0.center.mass"
  ),
  applies.to=c(
    "SPI and SPEI", "SPI and SPEI", "SPI and SPEI", "SPI and SPEI",
    "SPI only", "SPI and SPEI", "SPI and SPEI"
  ),
  values.and.default=c(
    "Numeric monthly zoo series; precipitation for SPI and P - PET for SPEI. The index must be Date, POSIXt, or yearmon and months must be consecutive.",
    "Any positive integer not exceeding the series length; no default. It controls the backward-looking memory and must be provided explicitly.",
    "list(type, shift); type is rectangular, triangular, circular, or gaussian and shift is 0,...,scale-1. Default list(type='rectangular', shift=0).",
    "NULL, a Date object, or a character string in YYYY-MM or YYYY-MM-DD format; both default to NULL, which fits the full record. They define the climatology used for parameter estimation, not the output period.",
    "Any finite non-negative number; default 0. Values strictly below it become zero before accumulation, allowing a trace-precipitation definition.",
    "TRUE or FALSE; defaults TRUE for SPI and FALSE for SPEI. TRUE represents exact zeros by a point mass plus a continuous positive distribution.",
    "TRUE or FALSE; default FALSE. With p0=TRUE, TRUE assigns zeros to the centre of their empirical probability mass rather than its upper edge."
  ),
  stringsAsFactors=FALSE
)
knitr::kable(data.arguments, row.names=FALSE)
```

The kernel encodes how memory is distributed within the selected scale.
`"rectangular"` gives all months equal weight. The other kernels progressively
change the relative influence of older observations, while `shift` moves the
weighting pattern within the backward-looking window. A non-rectangular kernel
therefore changes the scientific meaning of the index and should be justified
by the response dynamics being represented.

The reference period defines "normal". A short period produces uncertain tail
parameters because each calendar-month fit has approximately one observation
per reference year. A changing reference period also changes the climatology
against which drought is measured, so fixed and moving references answer
different questions.

### Distribution and parameter estimation

```{r theoretical-fitting-arguments, echo=FALSE}
fitting.arguments <- data.frame(
  argument=c(
    "distribution", "fit", "params", "start.fun", "start.fun.fix",
    "scaling", "..."
  ),
  applies.to=rep("SPI and SPEI", 7),
  values.and.default=c(
    "SPI: gamma (default), gumbel, logis, llogis, lnorm, norm, or weibull. SPEI: genlog (default), gev, norm, or pe3.",
    "max-lik (default), ub-pwm, or pp-pwm. It selects how each calendar month's distribution parameters are estimated.",
    "NULL (default), a parameter vector, an npar x 12 matrix, or an npar x nseries x 12 array. Supplied values define a fixed probability model and bypass fitting.",
    "NULL (default) or function(x, distr). It supplies starting parameters to maximum-likelihood optimisation for each month and series.",
    "TRUE or FALSE; default FALSE. TRUE substitutes starting estimates when optimisation fails; this improves continuity but does not constitute a converged MLE.",
    "sd (default), no, or max. Accumulated data are divided by their standard deviation, left unchanged, or divided by their maximum before fitting to improve numerical conditioning.",
    "Optional maximum-likelihood controls, principally mledist.par, such as optimiser, bounds, fixed parameters, and optimiser-specific controls."
  ),
  stringsAsFactors=FALSE
)
knitr::kable(fitting.arguments, row.names=FALSE)
```

`params` is appropriate when parameters have been estimated externally, when a
common climatology must be applied consistently, or when an operational system
must not refit retrospectively. Supplied parameters must use the parameter
order and convention documented in `?spi`; they apply to unscaled accumulated
data. Therefore, `fit`, `scaling`, `start.fun`, and `start.fun.fix` no longer
affect the result when `params` is supplied.

The accepted parameter names and order are:
`gamma=(shape, rate)`, `gumbel=(loc, scale)`,
`logis=(location, scale)`, `llogis=(shape, scale)`,
`lnorm=(meanlog, sdlog)`, `norm=(mean, sd)`,
`weibull=(shape, scale)`, `genlog=(shape, scale, location)`,
`gev=(loc, scale, shape)`, and `pe3=(shape, scale, location)`.
Scale and standard-deviation parameters must be positive; additional
distribution-specific constraints are checked by the function.

`scaling` is a numerical conditioning device, not a change to drought theory.
With internally estimated location/scale families, consistent rescaling should
not materially alter the standardised probabilities. It does matter to
optimisation stability and must not be mixed with parameter values calibrated
on a different scale.

### Missing values, bounds, and returned objects

```{r theoretical-output-arguments, echo=FALSE}
output.arguments <- data.frame(
  argument=c("sci.limit", "na.rm", "out.type", "verbose", "warn"),
  applies.to=rep("SPI and SPEI", 5),
  values.and.default=c(
    "Any non-negative number, including Inf (default). Finite values truncate the final index symmetrically to [-sci.limit, sci.limit].",
    "TRUE or FALSE; default FALSE. TRUE omits missing calibration values during fitting, but an accumulation window containing NA remains NA.",
    "zoo (default) or numeric. zoo preserves the time index and column names; numeric returns a vector or matrix.",
    "TRUE or FALSE; default FALSE. TRUE emits bracketed progress messages and does not change the calculation.",
    "TRUE or FALSE; default TRUE. FALSE suppresses fitting and transformation warnings but does not repair failed estimates."
  ),
  stringsAsFactors=FALSE
)
knitr::kable(output.arguments, row.names=FALSE)
```

Finite `sci.limit` values can prevent unstable extrapolation from producing
implausibly large magnitudes in a short calibration record (Stagge et al.,
2015). Truncation is a reporting safeguard, not an uncertainty estimate:
several observations can be assigned the same bound, and the true tail
probability remains uncertain.

# Part II: Practical use and package comparisons

The `spi()` and `spei()` functions calculate standardized drought indices from
complete monthly `zoo` series. Their production calculations use base R and do
not require the SCI, SPEI, or lmomco packages.

This practical part has two goals:

1. explain the main hydroTSM workflows for SPI and SPEI; and
2. document reproducible comparisons with SCI and SPEI at accumulation scales
   1, 3, 6, and 12.

The comparison section distinguishes equivalent computations from comparisons
that use different fitting methods or zero-frequency treatments. A high
correlation in a non-equivalent comparison is useful descriptive information,
but is not evidence of numerical reproduction under identical conditions.

## Differences from the SPEI and SCI packages

At the time of this benchmark, the CRAN versions were SPEI 1.8.1 and SCI
1.0-3. The comparison below concerns their documented public interfaces and the
hydroTSM 0.8-8 implementation, rather than suggesting that one design is
preferable for every workflow.

```{r implementation-differences, echo=FALSE}
implementation.table <- data.frame(
  aspect=c(
    "Primary interface", "Input and result", "Default calculation",
    "Distributions", "Parameter estimation", "Zero and tail controls",
    "PET calculation", "Fitted-model information", "Runtime dependencies"
  ),
  hydroTSM=c(
    "Direct spi() and spei() functions.",
    "Consecutive monthly zoo input; returns zoo or numeric values directly.",
    "scale is required; Gamma/p0 for SPI, genlog/no p0 for SPEI; max-lik fitting.",
    "All seven SPI and four SPEI candidates evaluated by Stagge et al. (2015).",
    "max-lik, ub-pwm, and pp-pwm; custom starts, fallback starts, or supplied parameters.",
    "zero.threshold for SPI; p0, centred zero mass, and symmetric sci.limit for either wrapper.",
    "Expects P - PET for spei(); PET must be computed separately.",
    "Returns the index, not a fitted-model object.",
    "Distribution fitting and optimisation use base R; no SCI, SPEI, lmomco, fitdistrplus, or evd dependency."
  ),
  SPEI=c(
    "Direct spi() and spei(); spi() is a wrapper around spei().",
    "Vector, matrix, data frame, or ts-like input; returns a spei object with fitted values and coefficients.",
    "scale is required; Gamma for SPI, log-Logistic for SPEI; ub-pwm fitting.",
    "Gamma, log-Logistic, and PearsonIII in the drought-index interface.",
    "ub-pwm, pp-pwm, and max-lik labels; available behaviour depends on distribution and data; supplied parameters supported.",
    "The spi()/spei() signatures do not expose zero.threshold, centred-zero-mass, or sci.limit controls.",
    "Package includes Thornthwaite, Hargreaves, and Penman PET functions.",
    "Returns coefficients, fitted values, settings, and optional input in a spei object.",
    "Imports lmomco, lmom, TLMoments, reshape, ggplot2, checkmate, zoo, and lubridate."
  ),
  SCI=c(
    "Separate fitSCI() and transformSCI() steps.",
    "Default methods operate on a monthly numeric univariate series plus first.mon; transformation returns numeric values.",
    "time.scale, distribution, and p0 are explicit; fitting is maximum likelihood.",
    "General SCI engine; bundled starting estimators cover nine named distributions and custom distributions can be supplied.",
    "Maximum likelihood with L-moment/moment starting estimates and custom start/fallback controls.",
    "p0, centred zero mass, scaling, and sci.limit are explicit; no zero.threshold argument.",
    "Expects the climate variable or P - PET to be prepared separately.",
    "fitSCI() returns monthly parameters and diagnostic flags for later transformation.",
    "Depends on fitdistrplus and lmomco; evd is suggested."
  ),
  check.names=FALSE,
  stringsAsFactors=FALSE
)
names(implementation.table)[3:4] <- c("SPEI 1.8.1", "SCI 1.0-3")
knitr::kable(implementation.table, row.names=FALSE)
```

The most important comparison consequence is the estimator default:
hydroTSM uses maximum likelihood, whereas SPEI defaults to unbiased PWM and SCI
fits by maximum likelihood. Equivalent numerical comparisons must therefore set
`fit` explicitly. The generalized-logistic model used by hydroTSM and SCI for
SPEI is functionally equivalent to the three-parameter log-logistic
formulation used by the SPEI package, but parameter names and signs must be
mapped consistently when parameters are supplied.

The current CRAN package descriptions and manuals are available at
<https://CRAN.R-project.org/package=SPEI> and
<https://CRAN.R-project.org/package=SCI>.

## Basic usage

```{r packages}
library(hydroTSM)
```

The main argument is a numeric monthly `zoo` object with a `Date`, `POSIXt`, or
`yearmon` index. The observations must cover consecutive months.

The following example uses monthly precipitation from one station in the
`EbroPPtsMonthly` dataset.

```{r input-data}
data(EbroPPtsMonthly)

pcp <- zoo::zoo(EbroPPtsMonthly$P9001, EbroPPtsMonthly$Date)
head(pcp)
```

### SPI

`scale` must be provided explicitly. With `scale=12`, `spi()` uses a Gamma
distribution, maximum-likelihood parameter estimation, and a mixed probability
at zero:

```{r spi-default}
spi12 <- spi(pcp, scale=12, warn=FALSE)
head(spi12, 15)
```

The accumulation scale is changed with `scale`. Trace precipitation can be
treated as zero before accumulation with `zero.threshold`:

```{r spi-scale}
spi1 <- spi(pcp, scale=1, zero.threshold=0.1, warn=FALSE)
spi3 <- spi(pcp, scale=3, zero.threshold=0.1, warn=FALSE)
spi6 <- spi(pcp, scale=6, zero.threshold=0.1, warn=FALSE)
```

The controls for low and zero precipitation have separate purposes:

- `zero.threshold` changes monthly values strictly below the threshold to zero;
- `p0=TRUE` uses a mixed distribution with a point probability at zero;
- `p0.center.mass=TRUE` uses the centered zero-mass estimator; and
- `sci.limit` bounds the final standardized index.

For example:

```{r spi-zero-treatment}
spi3.bounded <- spi(
  pcp, scale=3, zero.threshold=0.1,
  p0=TRUE, p0.center.mass=TRUE,
  sci.limit=3, warn=FALSE
)
range(spi3.bounded, na.rm=TRUE)
```

### SPEI

`spei()` expects monthly climatic water balance, normally precipitation minus
potential evapotranspiration. Its defaults select the generalized-logistic
distribution and do not use a zero-probability mixture.

For illustration, a deterministic seasonal PET series is constructed below:

```{r spei-data}
month.number <- as.integer(format(zoo::index(pcp), "%m"))
pet <- zoo::zoo(
  60 + 35 * sin(2 * pi * (month.number - 1) / 12),
  zoo::index(pcp)
)
water.balance <- pcp - pet

spei3 <- spei(water.balance, scale=3, warn=FALSE)
head(spei3)
```

SPI and SPEI retain the input time index. Initial values are `NA` when a complete
accumulation window is not yet available.

```{r indices-plot, fig.cap="Figure 1. Example three-month SPI and SPEI series."}
indices <- zoo::merge.zoo(SPI=spi3, SPEI=spei3)
plot(
  indices, plot.type="single", col=c("#1b9e77", "#d95f02"), lwd=1.2,
  xlab="Date", ylab="Standardized drought index"
)
abline(h=c(-2, -1, 0, 1, 2), col="grey85", lty=c(3, 3, 1, 3, 3))
legend(
  "topright", legend=colnames(indices),
  col=c("#1b9e77", "#d95f02"), lty=1, lwd=1.2, bty="n"
)
```

## Distributions and parameter estimation

The supported distributions are those evaluated for SPI and SPEI by Stagge
et al. (2015). The parameter column shows the required order when `params` is
supplied without names.

```{r spi-distributions, echo=FALSE}
spi.distribution.table <- data.frame(
  argument=c(
    "gamma", "gumbel", "logis", "llogis", "lnorm", "norm", "weibull"
  ),
  distribution=c(
    "Gamma", "Gumbel", "logistic", "log-logistic", "lognormal",
    "normal", "Weibull"
  ),
  parameters=c(
    "shape, rate", "loc, scale", "location, scale", "shape, scale",
    "meanlog, sdlog", "mean, sd", "shape, scale"
  )
)
knitr::kable(spi.distribution.table, row.names=FALSE)
```

```{r spei-distributions, echo=FALSE}
spei.distribution.table <- data.frame(
  argument=c("genlog", "gev", "norm", "pe3"),
  distribution=c(
    "generalized logistic", "GEV", "normal", "Pearson type III"
  ),
  parameters=c(
    "shape, scale, location", "loc, scale, shape", "mean, sd",
    "shape, scale, location"
  )
)
knitr::kable(spei.distribution.table, row.names=FALSE)
```

Three estimation methods are available:

- `fit="max-lik"`: maximum likelihood;
- `fit="ub-pwm"`: unbiased probability-weighted moments; and
- `fit="pp-pwm"`: plotting-position probability-weighted moments.

For example:

```{r fitting-methods}
spi3.mle <- spi(pcp, scale=3, fit="max-lik", warn=FALSE)
spi3.ub  <- spi(pcp, scale=3, fit="ub-pwm", warn=FALSE)
spi3.pp  <- spi(pcp, scale=3, fit="pp-pwm", warn=FALSE)
```

### Reference periods

Distribution parameters can be calibrated over a subperiod and then applied to
the complete series:

```{r reference-period}
spi3.reference <- spi(
  pcp, scale=3,
  ref.start="1943-01", ref.end="1948-12",
  warn=FALSE
)
```

Accumulation is performed before the reference-period subset is selected.
Therefore, an accumulated value at the start of the reference period can use
earlier observations from `x`. Character references can also include a day, for
example `"1943-01-01"`, or be supplied as `Date` objects. The day component is
only used to identify the calendar month.

### Supplied parameters

`params` bypasses parameter fitting. It accepts:

- a vector reused for every calendar month and series;
- an `npar` by 12 matrix of monthly parameters, reused for every series; or
- an `npar` by `nseries` by 12 array.

Named parameters are reordered automatically. Supplied parameters refer to the
unscaled accumulated data, so `scaling` is ignored.

Vector example: the same normal parameter set is reused for every calendar
month and every series.

```{r supplied-parameters-vector}
spei.fixed.vector <- spei(
  water.balance, scale=1, distribution="norm",
  params=c(mean=0, sd=40), warn=FALSE
)
head(spei.fixed.vector)
```

Matrix example: each calendar month receives its own parameter set, reused for
all series.

```{r supplied-parameters-matrix}
monthly.normal.params <- rbind(
  mean=seq(-10, 10, length.out=12),
  sd=rep(40, 12)
)
spei.fixed.monthly <- spei(
  water.balance, scale=1, distribution="norm",
  params=monthly.normal.params, warn=FALSE
)
head(spei.fixed.monthly)
```

Array example: each series and calendar month receives its own parameter set.

```{r supplied-parameters-array}
water.balance.two <- zoo::merge.zoo(
  stationA=water.balance,
  stationB=water.balance + 5
)
array.normal.params <- array(
  NA_real_, dim=c(2, 2, 12),
  dimnames=list(c("mean", "sd"), colnames(water.balance.two), month.abb)
)
array.normal.params["mean", "stationA", ] <- 0
array.normal.params["mean", "stationB", ] <- 5
array.normal.params["sd", , ] <- 40

spei.fixed.array <- spei(
  water.balance.two, scale=1, distribution="norm",
  params=array.normal.params, warn=FALSE
)
head(spei.fixed.array)
```

For SPI, `p0=TRUE` still estimates the monthly zero probability from the
reference data when continuous-distribution parameters are supplied.

### Custom maximum-likelihood starting values

`start.fun` follows the SCI interface and is called once for each calendar month
and series as `start.fun(x, distr)`. The default `NULL` uses hydroTSM's internal
base-R estimator.

```{r custom-start}
normal.start <- function(x, distr) {
  c(mean=mean(x), sd=stats::sd(x))
}

spei3.custom <- spei(
  water.balance, scale=3, distribution="norm",
  fit="max-lik", start.fun=normal.start,
  start.fun.fix=FALSE, warn=FALSE
)
```

When optimization fails, `start.fun.fix=FALSE` returns `NA` for the affected
month. Setting it to `TRUE` retains the initial values, which may be useful for
continuity but should not be interpreted as a converged maximum-likelihood fit.

### Progress messages

`verbose=TRUE` emits bracketed messages:

```{r verbose}
invisible(spi(pcp, scale=3, fit="ub-pwm", verbose=TRUE, warn=FALSE))
```

## Comparison design

The objective is to compare the SPI and SPEI series computed by the current
hydroTSM implementation with series computed by the SPEI and SCI packages. The
comparisons are made independently for accumulation scales 1, 3, 6, and 12 and
for hydroTSM `fit="max-lik"`, `fit="pp-pwm"`, and `fit="ub-pwm"`.

The benchmark uses the Wichita monthly dataset distributed with SPEI. Climatic
water balance was calculated with Thornthwaite PET at latitude 37.6475 degrees
north. All implementations used a rectangular kernel and the full record.

The package versions and data period used to create the bundled benchmark are:

```{r comparison-data, echo=FALSE}
comparison.file <- system.file(
  "extdata", "spi_spei_comparison_values.csv.gz", package="hydroTSM"
)
metadata.file <- system.file(
  "extdata", "spi_spei_comparison_metadata.csv.gz", package="hydroTSM"
)

if (!nzchar(comparison.file))
  comparison.file <- file.path("..", "inst", "extdata",
                               "spi_spei_comparison_values.csv.gz")
if (!nzchar(metadata.file))
  metadata.file <- file.path("..", "inst", "extdata",
                             "spi_spei_comparison_metadata.csv.gz")

read.comparison.csv <- function(file, ...) {
  if (grepl("\\.gz$", file)) {
    con <- gzfile(file, open="rt")
    on.exit(close(con), add=TRUE)
    utils::read.csv(con, ...)
  } else {
    utils::read.csv(file, ...)
  }
} # 'read.comparison.csv' END

comparison.values <- read.comparison.csv(
  comparison.file, check.names=FALSE, na.strings=""
)
comparison.values$date <- as.Date(comparison.values$date)
comparison.metadata <- read.comparison.csv(
  metadata.file, stringsAsFactors=FALSE
)

knitr::kable(comparison.metadata, row.names=FALSE)
```

All tables and plots use hydroTSM as the first series. The benchmark uses:

- Gamma with mixed zero probability for SPI;
- generalized logistic/log-logistic for SPEI;
- `scaling="sd"` for hydroTSM and SCI;
- scales 1, 3, 6, and 12; and
- all three hydroTSM fitting methods.

SCI supplies the maximum-likelihood reference. It has no `fit` argument
corresponding to hydroTSM's PWM alternatives, so comparisons of SCI with
`ub-pwm` or `pp-pwm` are descriptive rather than equivalent.

For SPEI 1.8.1, the same `fit` argument was requested from the package.
`ub-pwm` produced finite results. In this dataset, the `max-lik` and `pp-pwm`
runs produced no finite index values, so those cases are reported as not
computable rather than assigned an artificial error statistic.

At scale 1, Wichita contains exact zero precipitation. The two packages treat
those zero values differently in this configuration, so the scale-1 SPI
comparison with SPEI is not fully equivalent. At scales 3, 6, and 12 there are
no zero accumulated totals and the `ub-pwm` Gamma comparison is equivalent.

### Accuracy measures

Metrics are calculated only where both series are finite:

- correlation measures linear agreement;
- RMSE is the root mean squared difference;
- MAE is the mean absolute difference; and
- maximum error is the largest absolute difference.

```{r comparison-functions, echo=FALSE}
scales <- c(1L, 3L, 6L, 12L)
fits <- c("max-lik", "pp-pwm", "ub-pwm")
indices.names <- c("SPI", "SPEI")

comparison.column <- function(implementation, index, fit, scale) {
  paste(implementation, index, fit, paste0("s", scale), sep="__")
}

get.values <- function(implementation, index, fit, scale) {
  comparison.values[[comparison.column(implementation, index, fit, scale)]]
}

comparison.metrics <- function(first, second) {
  use <- is.finite(first) & is.finite(second)
  n <- sum(use)

  data.frame(
    pairs=n,
    correlation=if (n > 1L) stats::cor(first[use], second[use])
                else NA_real_,
    RMSE=if (n > 0L) sqrt(mean((first[use] - second[use])^2))
         else NA_real_,
    MAE=if (n > 0L) mean(abs(first[use] - second[use]))
        else NA_real_,
    max.error=if (n > 0L) max(abs(first[use] - second[use]))
              else NA_real_,
    hydro.nonfinite=sum(!is.finite(first)),
    reference.nonfinite=sum(!is.finite(second))
  )
}

results <- list()
k <- 0L

for (index in indices.names) {
  for (fit in fits) {
    for (scale in scales) {
      hydro <- get.values("hydroTSM", index, fit, scale)
      hydro.distribution <- if (index == "SPI") "gamma" else "genlog"
      hydro.label <- paste("hydroTSM", index, hydro.distribution, fit,
                           sep="; ")

      for (reference in c("SCI", "SPEI")) {
        reference.fit <- if (reference == "SCI") "max-lik" else fit
        reference.values <- get.values(
          reference, index, reference.fit, scale
        )
        current <- comparison.metrics(hydro, reference.values)
        reference.distribution <- if (reference == "SCI") {
          hydro.distribution
        } else if (index == "SPI") {
            "Gamma"
          } else {
              "log-Logistic"
            }
        reference.label <- paste(reference, index, reference.distribution,
                                 reference.fit, sep="; ")

        status <- if (reference == "SCI") {
          if (fit == "max-lik") "Fair" else "Different fitting methods"
        } else if (sum(is.finite(reference.values)) == 0L) {
          "No finite reference values"
        } else if (fit != "ub-pwm") {
          "Different fitting methods"
        } else if (index == "SPI" && scale == 1L) {
          "Different zero treatment"
        } else {
          "Fair"
        }

        k <- k + 1L
        results[[k]] <- data.frame(
          index=index,
          fit=fit,
          scale=scale,
          hydroTSM.result=hydro.label,
          reference=if (reference == "SCI")
                      "SCI (max-lik)" else paste("SPEI", fit),
          reference.result=reference.label,
          status=status,
          current,
          stringsAsFactors=FALSE
        )
      }
    }
  }
}

comparison.results <- do.call(rbind, results)

fair.sci.spi <- subset(
  comparison.results,
  index == "SPI" & fit == "max-lik" & reference == "SCI (max-lik)"
)
fair.sci.spei <- subset(
  comparison.results,
  index == "SPEI" & fit == "max-lik" & reference == "SCI (max-lik)"
)
fair.spei.spi <- subset(
  comparison.results,
  index == "SPI" & fit == "ub-pwm" & status == "Fair"
)
fair.spei.spei <- subset(
  comparison.results,
  index == "SPEI" & fit == "ub-pwm" & status == "Fair"
)
spi.scale1.spei <- subset(
  comparison.results,
  index == "SPI" & fit == "ub-pwm" & scale == 1 &
    grepl("^SPEI", reference)
)
```

## Independent comparison tables

Each table fixes one hydroTSM index and one hydroTSM fitting method, then
compares that hydroTSM series separately with SCI and SPEI at scales 1, 3, 6,
and 12. The `hydroTSM result` column identifies the series produced by the
current hydroTSM function. The `reference result` column identifies the external
package series used as the comparator. Both columns state the package, index,
distribution, and fitting method.

The `status` column identifies the conditions behind each row. `Fair` means the
distribution, fitting method, accumulation, reference period, scaling convention,
and zero treatment are equivalent for the comparison being made. Other status
values identify why the row is descriptive rather than a strict reproduction
test. The `nonfinite hydroTSM/reference` column reports the number of non-finite
values in each input series before pairwise metrics are calculated.

```{r comparison-tables, echo=FALSE, results="asis"}
for (index in indices.names) {
  for (fit in fits) {
    current <- subset(
      comparison.results,
      comparison.results$index == index & comparison.results$fit == fit
    )
    current$`nonfinite hydroTSM/reference` <- paste(
      current$hydro.nonfinite, current$reference.nonfinite, sep=" / "
    )
    current <- current[
      , c("scale", "hydroTSM.result", "reference.result", "status",
          "pairs", "correlation", "RMSE", "MAE", "max.error",
          "nonfinite hydroTSM/reference")
    ]
    names(current)[2:3] <- c("hydroTSM result", "reference result")

    cat("\n\n### ", index, ": `fit=\"", fit, "\"`\n\n", sep="")
    cat("Rows use the full Wichita record, a rectangular kernel, ",
        "`scaling=\"sd\"` for hydroTSM and SCI, and the distribution ",
        "named in the result columns.\n\n", sep="")
    print(knitr::kable(
      current,
      digits=c(0, NA, NA, NA, 0, 7, 7, 7, 7, NA),
      row.names=FALSE
    ))
    cat("\n")
  }
}
```

## Graphical summaries

Figures 2 to 5 show direct value-to-value agreement for the main fair
comparison families. A dashed one-to-one line represents exact agreement.

```{r plot-function, include=FALSE}
plot.comparison <- function(index, fit, reference, scale) {
  reference.fit <- if (reference == "SCI") "max-lik" else fit
  hydro <- get.values("hydroTSM", index, fit, scale)
  observed <- get.values(reference, index, reference.fit, scale)
  use <- is.finite(hydro) & is.finite(observed)

  if (!any(use)) {
    plot.new()
    title(main=paste("Scale", scale))
    text(0.5, 0.5, "No finite comparison pairs")
    return(invisible(NULL))
  }

  limits <- range(c(hydro[use], observed[use]))
  metrics <- comparison.metrics(hydro, observed)

  plot(
    observed[use], hydro[use],
    xlim=limits, ylim=limits, asp=1,
    pch=16, cex=0.55, col=grDevices::adjustcolor("#2166ac", 0.55),
    xlab=paste(reference, index),
    ylab=paste("hydroTSM", index),
    main=paste("Scale", scale)
  )
  abline(a=0, b=1, lty=2, col="#b2182b", lwd=1.2)
  legend(
    "topleft",
    legend=c(
      paste0("n = ", metrics$pairs),
      paste0("r = ", formatC(metrics$correlation, digits=6, format="f")),
      paste0("RMSE = ", formatC(metrics$RMSE, digits=3, format="g"))
    ),
    bty="n", cex=0.75
  )

  invisible(NULL)
}
```

```{r spi-sci-plot, echo=FALSE, fig.height=7.2, fig.cap="Figure 2. SPI: hydroTSM maximum likelihood versus SCI maximum likelihood."}
old.par <- par(mfrow=c(2, 2), mar=c(4.2, 4.2, 2.5, 1))
for (scale in scales)
  plot.comparison("SPI", "max-lik", "SCI", scale)
par(old.par)
```

```{r spei-sci-plot, echo=FALSE, fig.height=7.2, fig.cap="Figure 3. SPEI: hydroTSM maximum likelihood versus SCI maximum likelihood."}
old.par <- par(mfrow=c(2, 2), mar=c(4.2, 4.2, 2.5, 1))
for (scale in scales)
  plot.comparison("SPEI", "max-lik", "SCI", scale)
par(old.par)
```

```{r spi-spei-plot, echo=FALSE, fig.height=7.2, fig.cap="Figure 4. SPI: hydroTSM unbiased PWM versus the SPEI package. The scale-1 panel has different zero treatment."}
old.par <- par(mfrow=c(2, 2), mar=c(4.2, 4.2, 2.5, 1))
for (scale in scales)
  plot.comparison("SPI", "ub-pwm", "SPEI", scale)
par(old.par)
```

```{r spei-spei-plot, echo=FALSE, fig.height=7.2, fig.cap="Figure 5. SPEI: hydroTSM unbiased PWM versus the SPEI package."}
old.par <- par(mfrow=c(2, 2), mar=c(4.2, 4.2, 2.5, 1))
for (scale in scales)
  plot.comparison("SPEI", "ub-pwm", "SPEI", scale)
par(old.par)
```

## Interpretation

Under equivalent maximum-likelihood conditions, the largest RMSE against SCI
was `r formatC(max(fair.sci.spi$RMSE), digits=4, format="g")` for SPI and
`r formatC(max(fair.sci.spei$RMSE), digits=4, format="g")` for SPEI. All
correlations exceeded
`r formatC(min(c(fair.sci.spi$correlation, fair.sci.spei$correlation)), digits=8, format="f")`.

Under equivalent unbiased-PWM conditions, the largest RMSE against SPEI was
`r formatC(max(fair.spei.spi$RMSE), digits=4, format="g")` for SPI at scales
3, 6, and 12, and
`r formatC(max(fair.spei.spei$RMSE), digits=4, format="g")` for SPEI over all
four scales. These differences are at or close to floating-point precision.

The direct scale-1 SPI comparison with SPEI had correlation
`r formatC(spi.scale1.spei$correlation, digits=7, format="f")` and RMSE
`r formatC(spi.scale1.spei$RMSE, digits=5, format="f")`, but it is not a fair
zero-treatment comparison. It must not be combined with the equivalent
scale-3, scale-6, and scale-12 results when making a parity claim.

These results support two bounded conclusions for the tested data:

1. hydroTSM closely reproduces SCI when both use maximum likelihood, the same
   distributions, scaling, accumulation, and zero treatment; and
2. hydroTSM reproduces SPEI's unbiased-PWM results when the distribution and
   zero treatment are equivalent.

They do not establish universal equality for other datasets, fitting failures,
different reference periods, or intentionally different estimation methods.

## Reproducing the external-package benchmark

The benchmark values are bundled so this vignette can be built without making
SCI or SPEI package dependencies. A standalone script containing the complete
external-package computation is installed with hydroTSM:

```{r comparison-script, eval=FALSE}
comparison.script <- system.file(
  "extdata", "compare_spi_spei.R.txt", package="hydroTSM"
)
source(comparison.script)
```

SCI, SPEI, and zoo must be installed to run that validation script. By default,
it writes new compressed comparison and metadata CSV files to `tempdir()`. From
the hydroTSM source directory, explicit output files can be supplied with:

```text
Rscript inst/extdata/compare_spi_spei.R.txt values.csv.gz metadata.csv.gz
```

Because SCI and SPEI are comparison oracles rather than production
dependencies, updating either package should be followed by regenerating the
snapshot and reviewing all changes in finite-value coverage and accuracy.

# References {-}

McKee, T. B., Doesken, N. J., and Kleist, J. (1993). The relationship of
drought frequency and duration to time scales. *Proceedings of the 8th
Conference on Applied Climatology*, 179--184.

Guttman, N. B. (1998). Comparing the Palmer Drought Index and the
Standardized Precipitation Index. *Journal of the American Water Resources
Association*, 34, 113--121.
[doi:10.1111/j.1752-1688.1998.tb05964.x](https://doi.org/10.1111/j.1752-1688.1998.tb05964.x).

Hayes, M. J., Svoboda, M. D., Wilhite, D. A., and Vanyarkho, O. V. (1999).
Monitoring the 1996 drought using the Standardized Precipitation Index.
*Bulletin of the American Meteorological Society*, 80, 429--438.

Vicente-Serrano, S. M., Beguería, S., and López-Moreno, J. I. (2010).
A multiscalar drought index sensitive to global warming: the Standardized
Precipitation Evapotranspiration Index. *Journal of Climate*, 23, 1696--1718.
[doi:10.1175/2009JCLI2909.1](https://doi.org/10.1175/2009JCLI2909.1).

Beguería, S., Vicente-Serrano, S. M., Reig, F., and Latorre, B. (2014).
Standardized Precipitation Evapotranspiration Index (SPEI) revisited:
parameter fitting, evapotranspiration models, tools, datasets and drought
monitoring. *International Journal of Climatology*, 34, 3001--3023.
[doi:10.1002/joc.3887](https://doi.org/10.1002/joc.3887).

Greenwood, J. A., Landwehr, J. M., Matalas, N. C., and Wallis, J. R. (1979).
Probability weighted moments: definition and relation to parameters of
several distributions expressible in inverse form. *Water Resources
Research*, 15, 1049--1054.
[doi:10.1029/WR015i005p01049](https://doi.org/10.1029/WR015i005p01049).

Hosking, J. R. M. (1990). L-moments: analysis and estimation of distributions
using linear combinations of order statistics. *Journal of the Royal
Statistical Society: Series B*, 52, 105--124.
[doi:10.1111/j.2517-6161.1990.tb01775.x](https://doi.org/10.1111/j.2517-6161.1990.tb01775.x).

Stagge, J. H., Tallaksen, L. M., Gudmundsson, L., Van Loon, A. F., and Stahl,
K. (2015). Candidate distributions for climatological drought indices (SPI and
SPEI). *International Journal of Climatology*, 35, 4027--4040.
[doi:10.1002/joc.4267](https://doi.org/10.1002/joc.4267).

Pieper, P., Düsterhus, A., and Baehr, J. (2020). A universal Standardized
Precipitation Index candidate distribution function for observations and
simulations. *Hydrology and Earth System Sciences*, 24, 4541--4560.
[doi:10.5194/hess-24-4541-2020](https://doi.org/10.5194/hess-24-4541-2020).
